---
title: "Appendix A: Association Model"
author: "Bradley Hopkins, Matthew Hoctor"
date: "2/13/2021"
output:
  html_document:
    number_sections: yes
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
#library(tidyverse)
#library(ggplot2)
#library(ggthemes)
#library(gridExtra)
library(describedata)
#library(readxl)
#library(CarletonStats)
#library(pwr)
#library(BSDA)
#library(exact2x2)
#library(ppcor)
```

# Import the dataset

Import the maternal smoking dataset from the preliminary analysis:
```{r maternal smoking dataset, message = FALSE} 
CHDS <- read.csv("CHDS2.csv")
```

# Crude Model

The following R code creates the crude model and reports summary and ANOVA data:

```{r}
lm.mwbw <- lm(bwt ~ SMK_cat, data = CHDS)
summary(lm.mwbw)
anova(lm.mwbw)
```

# Full Model

The following R code creates the full model and reports summary and ANOVA data:

```{r}
lm.mwbw.full <- lm(bwt ~ SMK_cat + gestwks + age + mheight + mppwt + BMI, data = CHDS)
summary(lm.mwbw.full)
anova(lm.mwbw.full)
```

# Analysis of Effect Modification

## BMI

## Maternal Height

## Maternal Weight

## Maternal Age

## Gestational Age

# Analysis of Confounding

## BMI

## Maternal Height

## Maternal Weight

## Maternal Age

## Gestational Age

# Full Model Diagnostics

## Residual Analysis

## Normality Analysis

Due to the large sample size the Shapiro-Wilk test will not be performed.  The following R code creates a QQ Plot:



## Influential Outlier Detection

## Collinearity Assessment

# Mediation analysis

## BMI

## Maternal Weight

## Gestational Age
