---
title: "Preliminary Analysis"
author: "Bradley Hopkins, Matthew Hoctor"
date: "2/6/2021"
output:
  html_document:
    number_sections: yes
    theme: lumen
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(dplyr)
#library(readxl)
#library(tidyverse)
#library(ggplot2)
#library(CarletonStats)
#library(pwr)
#library(BSDA)
#library(exact2x2)
```

# Import the dataset

Import the maternal smoking dataset:
```{r maternal smomking dataset, message = FALSE}
CHDS <- read.csv("CHDS.csv")
```

# Descriptions of individual variables

## Birth Weight

## Gestational age

## Maternal Age

## Cigarettes smoked

## Maternal height

## Maternal pre-partum weight

# Generation of new variables of interest

## BMI

The following R Code creates a new variable for body mass index from the maternal weight and pre-partum height data:

```{r}
CHDS$BMI <- 703 * CHDS$mppwt / ((CHDS$mheight)^2)
```

## BMI category

The following R code creates a new variable for BMI category:

```{r}
CHDS$BMI_cat <- 
  ifelse(CHDS$BMI < 18.5, 0, 
  ifelse(CHDS$BMI < 25, 1,
  ifelse(CHDS$BMI < 30, 2,
  ifelse(CHDS$BMI < 35, 3,
  ifelse(CHDS$BMI < 40, 4, 
  5)))))
```

Note that 0 corresponds to underweight, 1 to normal weight, 2 to overweight, 3 to class I obese, 4 to class II obese, and 5 to class III obese.

## Smoking category

The following R code creates a new variable for smoking category.  Patients are categorized as 0, non-smokers; 1, light smokers 1-9 cigarettes/day; 2, moderate smokers, 10-19 cigarettes/day; or 3, heavy smokers 20+ cigarettes/day:

```{r}
CHDS$SMK_cat <-
  ifelse(CHDS$mnocig == 0, 0,
  ifelse(CHDS$mnocig < 10, 1,
  ifelse(CHDS$mnocig < 20, 2, 
  3)))
```

# Scatter plot matrix

The following R code creates a scatter plot matrix:

```{r}
pairs(~bwt+gestwks+age+mnocig+mheight+mppwt+BMI+BMI_cat+SMK_cat, data = CHDS, main = "Scatterplot Matrix")
```

# Check normality of response data

## Birth Weight

## Gestational Age

# References:

